Combining checkpointing and data compression\\ for high frequency seismic imaging

### 2018-10-02 paper/paper.tex
new keywords
Add images

\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{cuted}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\newcommand*\Laplace{\mathop{}\!\mathbin\bigtriangleup}
\begin{document}

@@

\section{Introduction}
The class of adjoint-based optimization problems has a very typical data access pattern where a
simulation is run ``forward" in simulation time, which produces data that is to be used by a
subsequent adjoint pass which is run backward in simulation time TODO CITE. Figure TODO
shows the typical data access pattern. This class consists of many different problems of interest in
science and engineering. One problem in this class is full-waveform inversion (FWI) where physical
experiments are carried out by propagating seismic waves through the earth's subsurface and the
solution to the optimization problem gives the most likely structure of the earth's subsurface. Figure
TODO shows the physical setup of the experiment that produces the data that this problem operates
on. The data access pattern from Figure TODO applies here as well, since every step during the
adjoint propagation requires data from the corresponding step from the forward propagation. In
particular for FWI, this data can reach tens of terabytes in size, which is beyond the memory
available on most hardware. While this paper is written to be focussed on FWI as the target problem, 
the analysis can be applied to any problem with a similar data-access pattern.

@@

@@

The values of $m(x)$ were derived from the SEAM Overthrust (TODO Cite SEAM) model which was over a grid
of $287 \times 881 \times 881$ points, including an absorbing layer of 40 points on each side. The grid spacing
was TODO in space and TODO in time. This was run for TODO timesteps, and the final field stored as
representative of a typical snapshot. Figure TODO shows this stored field that was used for the compression tests.

@@

\section{Compression algorithms}
\subsection{Lossless}
For our initial experiments, we used the python package \emph{blosc} (TODO CITE). It includes implementations for
6 different lossless compression algorithms, namely ZLIB, ZSTD, BLOSC, LZ4, LZ4HC and TODO (TODO cite all of them).
We did a parameter sweep over all available parameters to find the best-performing combinations. 
\subsection{Lossy}
Next we tried lossy compression using ZFP \cite{ratanaworabhan2006fast} (TODO verify citation).
We tried all three modes of ZFP, namely fixed-tolerance, fixed-precision and fixed-rate modes. Figure TODO shows the compression ratios
we saw for a variety of compression settings for the three modes. As expected, the highest compression efficiency was
observed in the fixed-tolerance mode albeit with highly unpredictable compression ratios. Figure TODO shows a sample
wavefield that was recovered after compression and decompression using TODO mode. Since, in our experiments, the time
taken to compress a checkpoint was in the same order of magnitude as the time taken to compute a timestep, we build a
performance model in the next section to evaluate the utility of compression in different scenarios of problem setups and test
platforms.

>>>>

\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{cuted}
\usepackage{tikz}
\usepackage{pgfplots}

\usetikzlibrary{calc,trees,positioning,arrows,chains,shapes.geometric,%
    decorations.pathreplacing,decorations.pathmorphing,shapes,%
    matrix,shapes.symbols}

\tikzstyle{line} = [draw, -, thick]
\tikzstyle{nodraw} = [draw, fill, circle, minimum width=0pt, inner sep=0pt]
\tikzstyle{box} = [line, rectangle, rounded corners, text centered]

\tikzset{
>=stealth',
  punktchain/.style={
    rectangle, 
    rounded corners, 
    draw=black, very thick,
    text width=2em, 
    minimum height=3em, 
    text centered, 
    on chain},
  line/.style={draw, thick, <-},
  element/.style={
    tape,
    top color=white,
    bottom color=blue!50!black!60!,
    minimum width=1em,
    draw=blue!40!black!90, very thick,
    text width=2em, 
    minimum height=3.5em, 
    text centered, 
    on chain},
  every join/.style={->, thick,shorten >=1pt},
  decoration={brace},
  tuborg/.style={decorate},
  tubnode/.style={midway, right=2pt},
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\newcommand*\Laplace{\mathop{}\!\mathbin\bigtriangleup}
\begin{document}

@@

\section{Introduction}
The class of adjoint-based optimization problems has a very typical data access pattern where a
simulation is run ``forward" in simulation time, which produces data that is to be used by a
subsequent adjoint pass which is run backward in simulation time TODO CITE. Figure \ref{fig:dataflow}
shows the typical data access pattern. This class consists of many different problems of interest in
science and engineering. One problem in this class is full-waveform inversion (FWI) where physical
experiments are carried out by propagating seismic waves through the earth's subsurface and the
solution to the optimization problem gives the most likely structure of the earth's subsurface. Figure
\ref{fig:offshore_survey} shows the physical setup of the experiment that produces the data that this problem operates
on. The data access pattern from Figure \ref{fig:dataflow} applies here as well, since every step during the
adjoint propagation requires data from the corresponding step from the forward propagation. In
particular for FWI, this data can reach tens of terabytes in size, which is beyond the memory
available on most hardware. While this paper is written to be focussed on FWI as the target problem, 
the analysis can be applied to any problem with a similar data-access pattern.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{images/survey-ship-diagram.png}
\end{center}
\caption{Illustration of an offshore seismic survey that collects the input data for FWI}
\label{fig:offshore_survey}
\end{figure}

@@

\begin{figure*}
\begin{center}
\input{dataflow}
\end{center}
\caption{The dataflow pattern that is typical of adjoint-based optimization problems}
\label{fig:dataflow}
\end{figure*}

@@

The values of $m(x)$ were derived from the SEAM Overthrust (TODO Cite SEAM) model which was over a grid
of $287 \times 881 \times 881$ points, including an absorbing layer of 40 points on each side. The grid spacing
was TODO in space and TODO in time. This was run for TODO timesteps, and the final field stored as
representative of a typical snapshot. Figure \ref{fig:uncompressed} shows this stored field that was used for the compression tests.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{images/uncompressed.png}
\end{center}
\caption{Cross-section of the wavefield used as a reference sample for compression and decompression}
\label{fig:uncompressed}
\end{figure}

@@

\section{Compression algorithms}
\subsection{Lossless}
For our initial experiments, we used the python package \emph{blosc} (TODO CITE). It includes implementations for
6 different lossless compression algorithms, namely ZLIB, ZSTD, BLOSC, LZ4, LZ4HC and TODO (TODO cite all of them).
We did a parameter sweep over all available parameters to find the best-performing combinations. 
\subsection{Lossy}
Next we tried lossy compression using ZFP \cite{ratanaworabhan2006fast} (TODO verify citation). To use ZFP from python, 
we first wrote a python wrapper over the reference implementation of ZFP, which is in C. We plan to release this wrapper
in the open source community shortly.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{images/decompressed-t-8.png}
\end{center}
\caption{Cross-section of the wavefield recovered after compression and decompression using the fixed-tolerance mode}
\label{fig:decompressed}
\end{figure}

We tried all three modes of ZFP, namely fixed-tolerance, fixed-precision and fixed-rate modes. Figure TODO shows the compression ratios
we saw for a variety of compression settings for the three modes. As expected, the highest compression efficiency was
observed in the fixed-tolerance mode albeit with highly unpredictable compression ratios. Figure \ref{fig:decompressed} shows a sample
wavefield that was recovered after compression and decompression using fixed-tolerance mode. Since, in our experiments, the time
taken to compress a checkpoint was in the same order of magnitude as the time taken to compute a timestep, we build a
performance model in the next section to evaluate the utility of compression in different scenarios of problem setups and test
platforms.
